---
title: "MATH 449 - Final Project"
author: "Dona Inayyah & Bryan Thorne"
date: "2023-05-18"
output:
  word_document: default
  pdf_document: default
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(MASS)
library(tibble)
library(Epi)
library(lattice)
library(DAAG)
library(knitr)
```

```{r config, include=FALSE}
# Folder containing Arrests.csv (override with the DATA_DIR environment variable)
DATA_DIR <- Sys.getenv("DATA_DIR", "data")
```

------------------------------------------------------------------------

#### State the problem and describe the data set

The data used for our final project, Arrests.csv, includes data on individuals in Toronto, Canada that were arrested for simple possession of small quantities of marijuana. The problem is to investigate the relationship between the release of an arrested individual, with a summons, and their race, sex, age, year of arrest, employment status, citizenship status, and number of checks (the number of times the arrested individual's name appeared in police databases).

This is an un-grouped dataset that consists of 5226 observations with 8 variables.

The variables are:

-   "released" (1=Yes, 0=No)

-   "colour" (1=White, 0=Black)

-   "sex" (1=Male, 0=Female)

-   "employed" (1=Yes, 0=No)

-   "citizen" (1=Yes, 0=No)

-   "year" (2002 - 1997)

-   "age" (12 - 66)

-   "checks" (0 - 6).

Variables "released", "color", "sex", "employed", and "citizens" are categorical whereas variables "year", "age" and "checks" are numerical. We will use "released" as the response variable, while the rest are explanatory variables.

#### Fit a logistic regression model with all predictors

```{r}
arrests = read.csv(file.path(DATA_DIR, "Arrests.csv"), header=TRUE)

# Transform response variable to binary
arrests$released = ifelse(arrests$released == "Yes", 1, 0)

# Transform categorical explanatory variables to binary 
arrests$sex = ifelse(arrests$sex == "Male", 1, 0)
arrests$employed = ifelse(arrests$employed == "Yes", 1, 0)
arrests$citizen = ifelse(arrests$citizen == "Yes", 1, 0)
arrests$colour = ifelse(arrests$colour == "White", 1, 0)

# Fit the logistic regression model with all predictors
fit0 = glm(released ~ colour + year + age + sex + employed + citizen + checks, 
           family = "binomial", data = arrests)
summary(fit0)
```

$logit[P(Y=1)]=0.974332+0.389109Co-0.004218Y+0.002236A+0.007317S+0.757302E+0.576519Ci-0.364101Ch$

-   Co represents the variable "colour".

-   Y represents the variable "year".

-   A represents the variable "age".

-   S represents the variable "sex".

-   E represents the variable "employed".

-   Ci represents the variable "citizen".

-   Ch represents the variable "checks".

#### Select the best subset of variables. Perform a diagnostic on the best model. Perform all possible inferences you can think about.

-   We can use **`stepAIC()`** to select the best subset of variables, as it uses forward selection and backward elimination.

```{r}
# Use stepAIC function to determine best subset of variables 
stepAIC(fit0)
```

```{r}
# Fit model with best subset 
fit1 = glm(released ~ colour + employed + citizen + checks, 
           family = "binomial", data = arrests)
summary(fit1)
```

By looking at the summary, we can see that race, employment status, citizenship status, and the number of checks are statistically significant at the 0.01 level. Therefore, we can infer that there is a strong association between these variables and the likelihood of an arrestee being released with a summons.

Before moving on, we should check to see if we have unbalanced data.

```{r}
table(arrests$released)
prop.released = 892/(892+4334)
cat("\n Proportion of not released:", prop.released)

I=which(arrests$released==1)
J=sample(I, 1000)
J1=which(arrests$released==0)
arrests1=rbind(arrests[J,],arrests[J1,])
```

By calculating the proportions of "released", we can see that we do have unbalanced data, where only 17% of the observation are not released. Therefore, we can create a subset of the data by sampling random 1000 observations from the "released" category (I) and combining them with all observations from the "not released" category (J1) to create a new, more balanced data-set called "**arrests1**".

Now we can fit a new model, with the same subset of variables used in **fit1**.

```{r}
fit2 = glm(released ~ colour + employed + citizen + checks, family = "binomial", data = arrests1)
(g=summary(fit2))

# Coefficients
alpha = g$coef[1,1]
beta1 = g$coef[2,1]
beta2 = g$coef[3,1]
beta3 = g$coef[4,1]
beta4 = g$coef[5,1]
```

Important note: Because our new, balanced **arrests1** data is combined with 1000 *random* observations from the "released" category, our estimate coefficient will vary after each run. However, this doesn't change its direction and the change in magnitude is very small.

-   Now we can carry out a Goodness-of-Fit Test with comparison to the null, to see if this is an adequate fit.

$H_0:$ All parameters in model $M$ not in the null model $M_0$ are zero

$H_1:$ At least one of them is not zero

```{r}
# Comparison with the null
fit_null = glm(released ~ 1, family = "binomial", data = arrests1)
anova(fit2, fit_null, test = "LRT")
```

Since we have a very small p-value that is less than a 0.001 level of significance, we have very strong evidence to reject the null hypothesis and conclude that our model is adequate.

We can now make inferences on our model.

-   Likelihood Ratio Confidence Intervals.

```{r}
# Confidence Interval
confint(fit2, level = 0.95)
```

-   Odds ratio

```{r}
# Multiplicative effect 
effect_co = exp(beta1)
effect_em = exp(beta2)
effect_ci = exp(beta3)
effect_ch = exp(beta4)

effect.df = data.frame(Predictor = c("Colour", "Employed", "citizen", "checks"), 
                       Effect = c(effect_co, effect_em, effect_ci, effect_ch))
effect.df
```

#### Use the new model to make predictions.

We can use the **`predict()`** function to make predictions.

```{r}
pihat = predict(fit2, type = "response")
```

We'll use four random observations as examples to further elaborate.

```{r}
Co=0; E=0; Ci=0; Ch=0
(exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))/
  (1+exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))
```

```{r}
Co=1; E=1; Ci=1; Ch=1
(exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))/
  (1+exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))
```

```{r}
Co=1; E=0; Ci=1; Ch=6
(exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))/
  (1+exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))
```

```{r}
Co=0; E=1; Ci=1; Ch=5
(exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))/
  (1+exp(alpha + beta1*Co + beta2*E + beta3*Ci + beta4*Ch))
```

#### Use different pi_0 as a cut-off point and create a confusion table.

There are two different ways to choose $\pi_0$ as a cut-off point.

-   $\pi_0=0.5$

```{r}
# pi0 = 0.5
y = as.numeric(arrests1$released > 0)
yhat = as.numeric(pihat > 0.50)
(confusion1 = addmargins(table(y, yhat), 2))

n11 = confusion1[1,1]
n12 = confusion1[1,2]
n1r = confusion1[1,3]
n21 = confusion1[2,1]
n22 = confusion1[2,2]
n2r = confusion1[2,3]
```

-   $\pi_0=\frac{(Y=1)}{n}$

```{r}
# pi0 = #(Y=1)/n 
pi_0 = n2r/(n1r+n2r)
y = as.numeric(arrests1$released > 0)
yhat = as.numeric(pihat > pi_0)
(confusion2 = addmargins(table(y, yhat), 2))

m11 = confusion2[1,1]
m12 = confusion2[1,2]
m1r = confusion2[1,3]
m21 = confusion2[2,1]
m22 = confusion2[2,2]
m2r = confusion2[2,3]
```

```{r}
# metrics for confusion1
sens1 = n22 / n2r
spec1 = n11 / n1r
acc1 = (n11 + n22) / (n1r + n2r)
err1 = (n12 + n21) / (n1r + n2r)

# metrics for confusion2
sens2 = m22 / m2r
spec2 = m11 / m1r
acc2 = (m11 + m22) / (m1r + m2r)
err2 = (m12 + m21) / (m1r + m2r)

metrics.df = data.frame(Matrix = c("Confusion 0", "Confusion 1"),
                        Sensitivity = c(sens1, sens2),
                        Specificity = c(spec1, spec2),
                        Accuracy = c(acc1, acc2),
                        Error = c(err1, err2))
metrics.df
```

#### Perform visualization of data and models.

-   Model different combinations of **fit2** to show the relationship between "released" & "checks"

```{r}
with(arrests1, {
  plot(jitter(checks, 5), released,
       xlim = c(-2, 8), ylim = c(0, 1),
       xlab = "Checks", ylab = "Released (Yes=1, No=0)")
})

# logit=alpha: curve(plogis(coef(fit2)[1]), col=1, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[5]*x), col=2, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[2] + coef(fit2)[3] + coef(fit2)[4] + coef(fit2)[5]*x), 
      col=3, add=TRUE, lwd=2)

curve(plogis(coef(fit2)[1] + coef(fit2)[4] + coef(fit2)[5]*x), col=4, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[3] + coef(fit2)[5]*x), col=5, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[2] + coef(fit2)[5]*x), col=6, add=TRUE, lwd=2)

curve(plogis(coef(fit2)[1] + coef(fit2)[2] + coef(fit2)[4] + coef(fit2)[5]*x), 
      col=7, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[3] + coef(fit2)[4] + coef(fit2)[5]*x), 
      col=8, add=TRUE, lwd=2)
curve(plogis(coef(fit2)[1] + coef(fit2)[2] + coef(fit2)[3] + coef(fit2)[5]*x), 
      col=9, add=TRUE, lwd=2)

lgnd = c("a + b4x",
         "a + b1 + b2 + b3 + b4x",
         "a + b1 + b4x",
         "a + b2 + b4x",
         "a + b3 + b4x",
         "a + b1 + b4 + b4x",
         "a + b2 + b4 + b4x", 
         "a + b3 + b4 + b4x")
legend("topright", lgnd, pch=19,
       col=1:9, text.col=1:41, cex = 0.55)
```

-   Box-plot of "checks"

```{r}
arrestyes=subset(arrests1, released==1)
arrestno=subset(arrests1, released==0)
boxplot(list(yes=arrestyes$checks, no=arrestno$checks))
```

Based on the box-plot, we can observe that the medians of the "yes" and "no" groups appear to be different. This suggests that there might be a difference in the central tendencies of the "checks" variable between the two groups. There is also an overlap between the "yes" and "no" groups, which indicates that both the groups have similar values for the number of checks.

-   Box-plot for "age"

```{r}
boxplot(list(yes=arrestyes$age, no=arrestno$age))
```

From the box-plot above, we can see that both groups have similar distributions, with nearly identical medians and both having outliers. This suggests that age may not be a strong predictor in determining whether an individual is being released with a summons.

#### Plot the ROC curve, find AUC, and the best cutoff point for classification.

-   ROC using $M_1$

```{r}
attach(arrests)

ROC(form=released ~ colour + employed + citizen + checks, plot="ROC")
```

-   ROC using $M_2$

```{r}
detach(arrests)
attach(arrests1)

ROC(form=released ~ colour + employed + citizen + checks, plot="ROC")
```

#### Perform LOOCV and k-fold cross-validation.

```{r}
# LOOCV
library(caret)
set.seed(123)

# Define training control
train.control = trainControl(method = "LOOCV")

model = released ~ colour + employed + citizen + checks

# Train the model
model = train(model, data = arrests1, method = "glm", trControl = train.control)
model
```

```{r}
# k-fold cross-validation.
cv.binary(fit1)
cv.binary(fit2)
```

#### Try the probit link and the identity links to model data.

```{r}
fit_probit = glm(released ~ colour + employed + citizen + checks, 
                 family = binomial(link = "probit"), data = arrests1)
summary(fit_probit)
```

```{r}
# fit_ident = glm(released ~ colour + employed + citizen + checks, 
                 #family = binomial(link = "identity"), data = arrests1)
#summary(fit_ident)
```

Using arrests1, we are not able to use the identity link. this could be due to a lack of convergence or co-linearity. Therefore, between the two, the probit link is better for this data.
